### 2025-03-13 paper/quarto_ecml/paper.tex
Review discussion

\textbf{\emph{CT is defined only for classification settings.}} Our
formulation relies on the distinction between non-target class(es)
\(y^{-}\) and target class(es) \(y^{+}\) to generate counterfactuals
through Equation~\ref{eq-obj}. While \(y^{-}\) and \(y^{+}\) can be
arbitrarily defined by the user, CT requires the output space
\(\mathcal{Y}\) to be discrete. Thus, it applies to binary and
multi-class classification but it is not well-defined for other ML tasks
where the change in outcome with respect to a decision threshold
\(\tau\) cannot be readily quantified. In fact, this is a common
restriction in research on CEs and AR that predominantly focuses on
classification models. Although other settings have attracted some
interest \cite{spooner2021counterfactual,zhao2023counterfactual}, there
is still no consensus on what constitutes a counterfactual in such
settings.

\textbf{\emph{CT is subject to training instabilities.}} Joint
energy-based models are susceptible to instabilities during training
\cite{grathwohl2020your} and even though we depart from the SGLD-based
sampling, we still encounter major variability in the outcomes. CT is
exposed to two potential sources of instabilities: (1) the energy-based
contrastive divergence term in Equation~\ref{eq-div}, and (2) the
underlying counterfactual explainers. For example,
\cite{altmeyer2023faithful} recognize this to be a challenge for
\emph{ECCCo} and so it may have downstream impacts on our proposed
method. Still, we find that training instabilities can be successfully
mitigated by regularizing energy (\(\lambda_{\text{reg}}\)), generating
a sufficiently large number of counterfactuals during each training
epoch and including only mature counterfactuals for contrastive
divergence.

\textbf{\emph{CT is sensitive to hyperparameter selection.}} As
discussed in Section~\ref{sec-hyperparameters}, our method benefits from
tuning certain key hyperparameters. In this work, we have relied
exclusively on grid search for this task. Future work on CT could
benefit from investigating more sophisticated approaches towards
hyperparameter tuning. Notably, CT is iterative which makes a variety of
methods applicable, including Bayesian \cite{snoek2012practical} or
gradient-based \cite{franceschi2017forward} optimization.

@@

\textbf{\emph{CT increases the training time of models.}} Counterfactual
training promotes explainability through CEs and robustness through AEs
at the cost of longer training times compared to conventional training
regimes. While higher numbers of iterations and counterfactuals per
iteration positively impact the quality of found solutions, they also
increase the required amount of computations. We find that relatively
small grids with 270 settings can take almost four hours for more
demanding datasets on a high-performance computing cluster with 34 2GB
CPUs (see details in \textbf{?@sec-hardware}). However, there are three
factors that attenuate the impact of this limitation. First, CT provides
counterfactual explanations for the training samples essentially for
free, which may be beneficial in many ADM systems. Second, we find that
CT can retain its value when used as a ``fine-tuning'' training regime
for conventionally-trained models. Third, in principle, CT yields itself
to parallel execution, which we have leveraged for our own experiments.

\textbf{\emph{Immutable features may have proxies.}} In
Proposition~\ref{prp-mtblty} we define an approach to protect immutable
features and thus increase the actionability of the generated
counterfactuals. However, this approach requires that model owners
define the mutability constraints for (all) features considered by the
model. Even with sufficient domain knowledge to protect all immutable
features---ones that cannot be changed at all and ones that cannot be
reasonably expected to change---there may exist proxies that are
theoretically mutable (and hence should not be protected) but preserve
enough information about the principals to counteract the protections.
As one example, consider the Adult dataset used in our experiments where
the mutable education status is a proxy for the immutable age, in that
the attainment of degrees is correlated with age. Delineating
actionability is a major undecided challenge in the AR literature
\cite{venkatasubramanian2020philosophical} impacting the capacity of CT
to increase the explainability of the model.

\textbf{\emph{Interventions on features may have downstream impacts on
fairness.}} Related to the point above, we provide a tool that allows
practitioners to modify the sensitivity of a model with respect to
certain features, which may have implication for the fair and equitable
treatment of individuals subject to automated decisions. As protecting a
set of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider again the
Adult dataset where features such as workclass or education may be more
difficult to change for underpriviledged groups. When applied
irresponsibly, counterfactual training could result in an unfairly
assigned burden of recourse \cite{sharma2020certifai}, threatening the
equality of opportunity in the system \cite{bell2024fairness} and
potentially reinforcing social segregation \cite{gao2023impact}. Still,
as the referenced publications indicate, such phenomena are not specific
to CT; all types of ADM solutions without strong external protections
have been recognized to promote harmful power dynamics
\cite{maas2023machine}.

@@

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

>>>>

\textbf{\emph{CT is defined only for classification settings.}} Our
formulation relies on the distinction between non-target class(es)
\(y^{-}\) and target class(es) \(y^{+}\) to generate counterfactuals
through Equation~\ref{eq-obj}. While \(y^{-}\) and \(y^{+}\) can be
arbitrarily defined, CT requires the output space \(\mathcal{Y}\) to be
discrete. Thus, it does not apply to ML tasks where the change in
outcome cannot be readily quantified. Focus on classification models is
a common restriction in research on CEs and AR. Other settings have
attracted some interest (e.g., regression in
\cite{spooner2021counterfactual,zhao2023counterfactual}), but there is
little consensus how to robustly extend the notion of counterfactuals.

\textbf{\emph{CT is subject to training instabilities.}} Joint
energy-based models are susceptible to instabilities during training
\cite{grathwohl2020your} and even though we depart from the SGLD-based
sampling, we still encounter major variability in the outcomes. CT is
exposed to two potential sources of instabilities: (1) the energy-based
contrastive divergence term in Equation~\ref{eq-div}, and (2) the
underlying counterfactual explainers. For example,
\cite{altmeyer2023faithful} recognize this to be a challenge for
\emph{ECCCo} and so it may have downstream impacts on our proposed
method. Still, we find that training instabilities can be successfully
mitigated by regularizing energy (\(\lambda_{\text{reg}}\)), generating
a sufficiently large number of counterfactuals during each training
epoch, and including only mature counterfactuals for contrastive
divergence.

\textbf{\emph{CT is sensitive to hyperparameter selection.}} Our method
benefits from tuning certain key hyperparameters (see
Section~\ref{sec-hyperparameters}). In this work, we have relied
exclusively on grid search for this task. Future work on CT could
benefit from investigating more sophisticated approaches towards
hyperparameter tuning. Notably, CT is iterative which makes a variety of
methods applicable, including Bayesian \cite{snoek2012practical} or
gradient-based \cite{franceschi2017forward} optimization.

@@

\textbf{\emph{CT increases the training time of models.}} Counterfactual
training promotes explainability through CEs and robustness through AEs
at the cost of longer training times compared to conventional training
regimes. While higher numbers of iterations and counterfactuals per
iteration positively impact the quality of found solutions, they also
increase the required amount of computations. We find that relatively
small grids with 270 settings can take almost four hours for more
demanding datasets on a high-performance computing cluster with 34 2GB
CPUs (see details in \textbf{?@sec-hardware}). However, there are three
factors that attenuate the impact of this limitation. First, CT provides
CEs for the training samples for free, which may be beneficial in many
ADM systems. Second, we find that CT can retain its value when used as a
``fine-tuning'' training regime for conventionally-trained models.
Third, in principle, CT yields itself to parallel execution, which we
have leveraged for our own experiments.

\textbf{\emph{Immutable features may have proxies.}} We propose an
approach to protect immutable features and thus increase the
actionability of the generated CEs. However, it requires that model
owners define the mutability constraints for (all) features considered
by the model. Even with sufficient domain knowledge to protect all
immutable features, there may exist proxies that are theoretically
mutable (and hence should not be protected) but preserve enough
information about the principals to hinder the protections. As an
example, consider the Adult dataset used in our experiments where the
mutable education status is a proxy for the immutable age, in that the
attainment of degrees is correlated with age. Delineating actionability
is a major undecided challenge in the AR literature
\cite{venkatasubramanian2020philosophical} impacting the capacity of CT
to increase the explainability of the model.

\textbf{\emph{Interventions on features may impact fairness
downstream.}} Related to the point above, we provide a tool that allows
practitioners to modify the sensitivity of a model with respect to
certain features, which may have implication for the fair and equitable
treatment of individuals subject to automated decisions. As protecting a
set of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider again the
Adult dataset where features such as workclass or education may be more
difficult to change for underpriviledged groups. When applied
irresponsibly, counterfactual training could result in an unfairly
assigned burden of recourse \cite{sharma2020certifai}, threatening the
equality of opportunity in the system \cite{bell2024fairness} and
potentially reinforcing social segregation \cite{gao2023impact}. Still,
as the referenced publications indicate, such phenomena are not specific
to CT; all types of ADM solutions without strong external protections
have been recognized to promote harmful power dynamics
\cite{maas2023machine}.

@@

### 2025-03-13 paper/sections/discussion.qmd
Review discussion

***CT is defined only for classification settings.*** Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined by the user, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it applies to binary and multi-class classification but it is not well-defined for other ML tasks where the change in outcome with respect to a decision threshold $\tau$ cannot be readily quantified. In fact, this is a common restriction in research on CEs and AR that predominantly focuses on classification models. Although other settings have attracted some interest [e.g., regression in @spooner2021counterfactual; @zhao2023counterfactual], there is still no consensus on what constitutes a counterfactual in such settings.

***CT is subject to training instabilities.*** Joint energy-based models are susceptible to instabilities during training [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter major variability in the outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. For example, @altmeyer2023faithful recognize this to be a challenge for *ECCCo* and so it may have downstream impacts on our proposed method. Still, we find that training instabilities can be successfully mitigated by regularizing energy ($\lambda_{\text{reg}}$), generating a sufficiently large number of counterfactuals during each training epoch and including only mature counterfactuals for contrastive divergence.

***CT is sensitive to hyperparameter selection.*** As discussed in @sec-hyperparameters, our method benefits from tuning certain key hyperparameters. In this work, we have relied exclusively on grid search for this task. Future work on CT could benefit from investigating more sophisticated approaches towards hyperparameter tuning. Notably, CT is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.

@@

***CT increases the training time of models.*** Counterfactual training promotes explainability through CEs and robustness through AEs at the cost of longer training times compared to conventional training regimes. While higher numbers of iterations and counterfactuals per iteration positively impact the quality of found solutions, they also increase the required amount of computations. We find that relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs (see details in @sec-hardware). However, there are three factors that attenuate the impact of this limitation. First, CT provides counterfactual explanations for the training samples essentially for free, which may be beneficial in many ADM systems. Second, we find that CT can retain its value when used as a "fine-tuning" training regime for conventionally-trained models. Third, in principle, CT yields itself to parallel execution, which we have leveraged for our own experiments.

***Immutable features may have proxies.*** In @prp-mtblty we define an approach to protect immutable features and thus increase the actionability of the generated counterfactuals. However, this approach requires that model owners define the mutability constraints for (all) features considered by the model. Even with sufficient domain knowledge to protect all immutable features---ones that cannot be changed at all and ones that cannot be reasonably expected to change---there may exist proxies that are theoretically mutable (and hence should not be protected) but preserve enough information about the principals to counteract the protections. As one example, consider the Adult dataset used in our experiments where the mutable education status is a proxy for the immutable age, in that the attainment of degrees is correlated with age. Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to increase the explainability of the model.

***Interventions on features may have downstream impacts on fairness.***  Related to the point above, we provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of individuals subject to automated decisions. As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider again the Adult dataset where features such as workclass or education may be more difficult to change for underpriviledged groups. When applied irresponsibly, counterfactual training could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness] and potentially reinforcing social segregation [@gao2023impact]. Still, as the referenced publications indicate, such phenomena are not specific to CT; all types of ADM solutions without strong external protections have been recognized to promote harmful power dynamics [@maas2023machine].

>>>>

***CT is defined only for classification settings.*** Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily quantified. Focus on classification models is a common restriction in research on CEs and AR. Other settings have attracted some interest (e.g., regression in [@spooner2021counterfactual; @zhao2023counterfactual]), but there is little consensus how to robustly extend the notion of counterfactuals.

***CT is subject to training instabilities.*** Joint energy-based models are susceptible to instabilities during training [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter major variability in the outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. For example, @altmeyer2023faithful recognize this to be a challenge for *ECCCo* and so it may have downstream impacts on our proposed method. Still, we find that training instabilities can be successfully mitigated by regularizing energy ($\lambda_{\text{reg}}$), generating a sufficiently large number of counterfactuals during each training epoch, and including only mature counterfactuals for contrastive divergence.

***CT is sensitive to hyperparameter selection.*** Our method benefits from tuning certain key hyperparameters (see @sec-hyperparameters). In this work, we have relied exclusively on grid search for this task. Future work on CT could benefit from investigating more sophisticated approaches towards hyperparameter tuning. Notably, CT is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.

@@

***CT increases the training time of models.*** Counterfactual training promotes explainability through CEs and robustness through AEs at the cost of longer training times compared to conventional training regimes. While higher numbers of iterations and counterfactuals per iteration positively impact the quality of found solutions, they also increase the required amount of computations. We find that relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs (see details in @sec-hardware). However, there are three factors that attenuate the impact of this limitation. First, CT provides CEs for the training samples for free, which may be beneficial in many ADM systems. Second, we find that CT can retain its value when used as a "fine-tuning" training regime for conventionally-trained models. Third, in principle, CT yields itself to parallel execution, which we have leveraged for our own experiments.

***Immutable features may have proxies.*** We propose an approach to protect immutable features and thus increase the actionability of the generated CEs. However, it requires that model owners define the mutability constraints for (all) features considered by the model. Even with sufficient domain knowledge to protect all immutable features, there may exist proxies that are theoretically mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections. As an example, consider the Adult dataset used in our experiments where the mutable education status is a proxy for the immutable age, in that the attainment of degrees is correlated with age. Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to increase the explainability of the model.

***Interventions on features may impact fairness downstream.***  Related to the point above, we provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of individuals subject to automated decisions. As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider again the Adult dataset where features such as workclass or education may be more difficult to change for underpriviledged groups. When applied irresponsibly, counterfactual training could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness] and potentially reinforcing social segregation [@gao2023impact]. Still, as the referenced publications indicate, such phenomena are not specific to CT; all types of ADM solutions without strong external protections have been recognized to promote harmful power dynamics [@maas2023machine].
